Storm surge modelling in the Strait of Georgia

### 2014-11-22 RoughWork/manuscript.tex
Text for Dec 2012 case. Modifcations to abstract and discussion

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas during a major wind storm on Dec 15, 2006. Generalll, surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. Another group of modellers studying a larger domain noted high temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. A finite volume model on an unstrcutured grid by \citet{yang2010multi} used a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. \citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia and noted that their semidiurnal amplitudes were generally too low. The passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling of this region.
%Add more about reproduction of tides in other models

@@

\citet{danard2003storm} discuss the role of wind fetch, the distance over which wind acts on the surface of a water basin, in setting up seiche motion in lakes and its contribution to storm surges. As wind blows over the surface of a lake, the water level becomes elevated on the downwind side and depressed on the upwind side in response to the wind-induced currents. The longer the fetch, the more significant the tilting of the water surface. These changes in water level can also contribute to storm surges.  During the high wind event on Nov 19 and Nov 20, the winds at Point Atkinson were directed to the west, explaining the lack of strong surge observed during this event since wind stresses would act to depress the water on the east side of the basin. In fact, the surge amplitude at Victoria on Nov 20 as 64.7 cm which is slightly higher than that observed at Point Atkinson (62.5 cm). Of the cases considered, this is the only example where the surge amplitude at Victoria was higher than that at Point Atkinson. In fact, this event observed a higher surge at Victoria than the Dec 2006 event, further highlighting the role that winds plays in setting up local variations in the surge. In this case, the model results do not reflect a higher surge at Victoria (46.1 cm versus 50.5 cm at Point Atkinson), perhaps due to an underestimate in the wind speed and some discrepancies in the wind direction of the atmospheric forcing.

@@

\subsection{High Resolution Atmospheric Forcing}
The above comparison between the modelled and observed winds at Point Atkinson suggests that the atmopsheric forcing product (CGRF) may not be adequate to produce accurate storm surge hindcasts since the wind speeds are substanitially lower than observed. CGRF is a global atmospheric reforecasting model with horizontal resolution approximately 33 km over the Salish Sea domain, which is problematic since the Strait of Georgia itself is only about 30 km wide. As such, there are only one or two grid cells that are entirely over water in our domain, resulting in overly damped wind speeds at grid points close to land such as Point Atkinson. To determine the effect of underresolved wind forcing on the storm surge hindcasts we compare simulations that have been forced with 1) the low resolution CGRF model and 2) a high resolution operational model from Environment Canada.

%Where can I find a reference for the atmospheric model?

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2012_weather.pdf}
\caption{Comparison of observed and modelled winds during Dec 2012 storm surge simulation at Point Atkinson and Sandheads.}
\label{fig:dec2013_weather}
\end{figure}
%don't include direction?

@@

Hindcasts of four significant events in February 2006, November 2006, December 2006 and November 2009 were simulated. In general, the model performs well at reproducing the observed water level heights and surge amplitudes, with Willmott skill scores greater than 0.88 in all cases. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude ($>50$ \%), in agreement with previous studies in this region \citep{murty1995storm}. We have also found that the wind and atmospheric conditions can set up spatial differences in the surge amplitude, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are usually higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important after the San Juan and Gulf Islands. The modelled surge amplitude increased by up to 10 \% at Point Atkinson when the tide-surge interaction was neglected.

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, sea surface height anomaly from Neah Bay will be used in future real time water level predictions during the storm surge season. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas.

Since the Strait of Georgia is a large, semi-enclosed body of water storm surges can behave and evolve differently than in other coastal regions. For instance, the complicated coastlines and local wind patterns result in important spatial variations in the surge amplitude. As such, the strong wind caused by a large storm are not the only factors to be considered in community preparedness for surge conditions. The geometry of the Strait and local wind patterns render communities on the mainland side of the southern Strait of Georgia more susceptible to surges. Additionally, commmunities such as Campbell River are at risk due to the long fetch and predominately southerly winds during the winter months.

@@

\section{Acknowledgements}\label{sec:ack}
We would like to thank Michael Foreman, Diane Masson, Luc Fillion Rich Pawlowicz, and Mark Halverson for providing data used in model set up and evaluation. This project is funded by the Marine Environmental Observation Prediction and Response Network (MEOPAR), a Network of Centres of Excellence of Canada.

>>>>

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. Comparisons between simulations with high resolution and low resolution atmospheric forcing further emphasize this point. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas during a major wind storm on Dec 15, 2006. Generally, surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
\end{abstract}

@@

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. Another group of modellers studying a larger domain noted high temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. A finite volume model on an unstrcutured grid by \citet{yang2010multi} used a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. \citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia and noted that their semidiurnal amplitudes were generally too low. The passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling of this region.

@@

\citet{danard2003storm} discuss the role of wind fetch, the distance over which wind acts on the surface of a water basin, in setting up seiche motion in lakes and its contribution to storm surges. As wind blows over the surface of a lake, the water level becomes elevated on the downwind side and depressed on the upwind side in response to the wind-induced currents. The longer the fetch, the more significant the tilting of the water surface. These changes in water level can also contribute to storm surges.  During the high wind event on Nov 19 and Nov 20, the winds at Point Atkinson were directed to the west, explaining the lack of strong surge observed during this event since wind stresses would act to depress the water on the east side of the basin. In fact, the surge amplitude at Victoria on Nov 20 was 64.7 cm which is slightly higher than that observed at Point Atkinson (62.5 cm). Of the cases considered so far, this is the only example where the surge amplitude at Victoria was higher than that at Point Atkinson. In fact, this event observed a higher surge at Victoria than the Dec 2006 event, further highlighting the role that winds plays in setting up local variations in the surge. In this case, the model results do not reflect a higher surge at Victoria (46.1 cm versus 50.5 cm at Point Atkinson), perhaps due to an underestimate in the wind speed and some discrepancies in the wind direction of the atmospheric forcing.

@@

\subsection{High Resolution Atmospheric Forcing}
The above comparison between the modelled and observed winds at Point Atkinson suggests that the atmopsheric forcing product (CGRF) may not be adequate to produce accurate storm surge hindcasts since the wind speeds are substanitially lower than observed. CGRF is a global atmospheric reforecasting model with horizontal resolution approximately 33 km over the Salish Sea domain, which is problematic since the Strait of Georgia itself is only about 30 km wide. As such, there are only one or two grid cells that are entirely over water in our domain, resulting in overly damped wind speeds at grid points close to land such as Point Atkinson. To determine the effect of underresolved wind forcing on the storm surge hindcasts we compare simulations that have been forced with 1) the low resolution CGRF model and 2) a high resolution (2.5 km) operational model from Environment Canada.  These simulations are performed for a storm surge that occured on Dec 17, 2012 and cover the time period Dec 14-19, 2012. The high resolution product is not available for the hindcasts already discussed.

First, a comparison of the wind speed and direction for the two models and observations is provided in Figure \ref{fig:dec2012_weather} at Point Atkinson and at Sandheads, a lighthouse station in the southern Strait of Georgia. For each of the models, the wind speed and direction at the closest grid point to the longitude/latitude of Point Atkinson and Sandheads is used for comparison. Although the low resolution CGRF model has a grid spacing of 33 km, two grid points align closely with these locations. At Point Atkinson, the high resolution model is in very good agreement with observations for both the wind speed and direction, however, the low resolution CGRF model significantly underestimates the wind speed during high wind periods on Dec 15, 2012 and Dec 17, 2012. At Sandheads, both the low resolution and high resolution models have good agreement with observations. The low resolution model, surprsingly, has a very good match likely because this grid cell happens to be mostly over water. Although not shown here, both models have a similar sea level atmospheric pressure at these two locations.

%Where can I find a reference for the high res atmospheric model?

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2012_weather.pdf}
\caption{Comparison of observed and modelled winds during Dec 2012 storm surge simulation at Point Atkinson (upper) and Sandheads (lower). Observed winds shown in solid black, low resolution CGRF model in dashed black and high resolution model in gray. }
\label{fig:dec2012_weather}
\end{figure}
%don't include direction?

Modelled and observed residuals for this simulation are displayed in Figure \ref{fig:dec2012_surge} where it is apparent that both of the models agree well with the observations. Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmopsheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occured later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly underpredicts the wind speed at some locations, it performs comparable to forcing with high resolution prdocuts. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

@@

Hindcasts of four significant events in February 2006, November 2006, December 2006 and November 2009 were simulated. In general, the model performs well at reproducing the observed water level heights and surge amplitudes, with Willmott skill scores greater than 0.88 in all cases. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude ($>50$\%), in agreement with previous studies in this region \citep{murty1995storm}. We have also found that the wind and atmospheric conditions can set up spatial differences in the surge amplitude, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are typically higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important after the San Juan and Gulf Islands. The modelled surge amplitude increased by about 10\% at Point Atkinson when the tide-surge interaction was neglected.

Comparisons between high resolution and low resolution atmospheric forcing further suggest that the local winds are not the leading order factor in setting up large surge amplitudes. The low resolution model significantly underestimated wind speeds in certain areas of the domain, however, it performed just as accurately as the high resolution product.  Although the high resolution model delivers fine scale detail and more accurate local wind patterns, it did not provide any improvement to the surge hindcast. Since both of these products provide the same level of accuracy in storm surge hindcasts, in the future we will use the high resolution atmospheric model to produce real-time storm surge predictions, even though most of out testing was performed with the low resolution model.

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, the sea surface height anomaly from Neah Bay will be used in future real time water level predictions during the storm surge season. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas.

Since the Strait of Georgia is a large, semi-enclosed body of water storm surges can behave and evolve differently than in other coastal regions. For instance, the complicated coastlines and local wind patterns result in important spatial variations in the surge amplitude. As such, the strong wind caused by a large storm are is not the only factor to be considered in community preparedness for surge conditions. The geometry of the Strait and local wind patterns render communities on the mainland side of the southern Strait of Georgia more susceptible to surges. Additionally, commmunities such as Campbell River are at risk due to the long fetch and predominately southerly winds during the winter months.

@@

\section{Acknowledgements}\label{sec:ack}
We would like to thank Michael Foreman, Diane Masson, Luc Fillion, Rich Pawlowicz, and Mark Halverson for providing data used in model set up and evaluation. This project is funded by the Marine Environmental Observation Prediction and Response Network (MEOPAR), a Network of Centres of Excellence of Canada.
